---
title: "Final project of Data Science Lab"
subtitle: <i>di Beatrice Fumagalli, Matteo Porcino e Pierluigi Tagliabue</i>
output: html_notebook
---

### <span style="color:DODGERBLUE"> Import libraries </span>
```{r, warning=FALSE}
library(dplyr)
library(kableExtra)
library(kohonen)
library(corrplot)
library(toaster)
library(NbClust)
library(Hmisc)
```
#------------------------------------------------------------------------------------------------------

### <span style="color:DODGERBLUE"> Data Manipulation </span>

#### Importing data and showing the structure of the dataset after removing the useless columns

```{r,  warning=FALSE}
db <- read.csv("D:\\Documents\\UNIMIB\\Secondo anno\\DS Lab\\DatasetClientClustering.csv", sep = ",", header = TRUE)
db$Features <- NULL
db$Descrizione <- NULL
db$X <- NULL 
db$X.1 <- NULL
db$X.2 <- NULL
db$X.3 <- NULL
db$X.4 <- NULL
db$X.5 <- NULL

head((db)) 
```
#------------------------------------------------------------------------------------------------------

```{r}
str(db)
```
#------------------------------------------------------------------------------------------------------

#### Assign string 'NA' (as in Napoli) to dataframe's row with Prov value as NA (recognised as NaN)

```{r}
db_na <- na.omit(db)
```
#------------------------------------------------------------------------------------------------------

#### Set Prov variable as factor and then numeric

```{r}
db_na$Prov <- factor(db_na$Prov)
db_na$Prov <- as.numeric(db_na$Prov)
```
#------------------------------------------------------------------------------------------------------

#### Remove ClientID column

```{r}
db_na$ClientID <- NULL

db_na$PanicMood <- as.numeric(db_na$PanicMood)
db_na$PanicMood[db_na$PanicMood == -1] <- 1
#db_na$PanicMood <- as.factor(db_na$PanicMood)
#db_na$NoTrustInBanks=as.factor(db_na$NoTrustInBanks)
```
#------------------------------------------------------------------------------------------------------

#### Attributes have different units of measure, so they need to be standardized

```{r}
data_scale <- as.data.frame(scale(db_na))
head(data_scale)
```
#------------------------------------------------------------------------------------------------------

# <span style="color:DODGERBLUE"> Matrice di correlazione </span>

```{r}
library(reshape2)
library(ggplot2)
cormat <- round(cor(data_scale), 2)
melted_cormat <- melt(cormat)
#ggplot(data = melted_cormat, aes(x=Var1, y=Var2, fill=value)) + geom_tile()

get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    return(cormat)
}

reorder_cormat <- function(cormat){
# Use correlation between variables as distance
dd <- as.dist((1-cormat)/2)
hc <- hclust(dd)
cormat <-cormat[hc$order, hc$order]
}

cormat <- reorder_cormat(cormat)
upper_tri <- get_upper_tri(cormat)
# Melt the correlation matrix
melted_cormat_up <- melt(upper_tri, na.rm = TRUE)
# Create a ggheatmap
ggheatmap <- ggplot(melted_cormat_up, aes(Var2, Var1, fill = value))+
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
    name="Pearson\nCorrelation") +
  theme_minimal()+ # minimal theme
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 10, hjust = 1))+
 coord_fixed()

graph <- ggheatmap + 
geom_text(aes(Var2, Var1, label = value), color = "black", size = 2) +
theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  #panel.grid.major = element_blank(),
  panel.border = element_blank(),
  panel.background = element_blank(),
  axis.ticks = element_blank(),
  legend.justification = c(1, 0),
  legend.position = c(0.6, 0.7),
  legend.direction = "horizontal")+
  guides(fill = guide_colorbar(barwidth = 7, barheight = 1,
                title.position = "top", title.hjust = 0.5))

ggsave(graph, file="Corr_graph.svg")
graph
```
#------------------------------------------------------------------------------------------------------

# <span style="color:DODGERBLUE"> Cluster 1- Variabili correlate positivamente</span>

```{r}
var_corr_pos <- data_scale[, c(10, 11, 15)]
```

```{r}
var_corr_pos 
```

```{r}
library(cluster)
library(dplyr)

km.out <- list()
sil.out <- list()
x <- vector()
y <- vector()
minClust <- 2     
maxClust <- 10



for (centr in minClust:maxClust) {
  i <- centr-(minClust-1) 
  km.out[i] <- list(kmeans(var_corr_pos, centers = centr, nstart = 50,iter.max = 50))
  sil.out[i] <- list(silhouette(km.out[[i]][[1]], dist(var_corr_pos)))
  x[i] = centr  # value of k
  y[i] = summary(sil.out[[i]])[[4]]  # Silhouette average width
}


library(ggplot2)
ggplot(data = data.frame(x, y), aes(x, y)) + 
  geom_point(size=3) + 
  geom_line() +
  xlab("Number of Cluster Centers") +
  ylab("Silhouette Average Width") +
  ggtitle("Silhouette Average Width as Cluster Center Varies")

ks <- 2:10 # number of clusters we want to try
ssw <- numeric(length(ks)) # vector for the ss_within
for (i in seq_along(ks)) {
  ssw[i] <- kmeans(var_corr_pos, ks[i],iter.max = 50)$tot.withinss
}

plot(x = ks, y = ssw, type = "l",
     xlab = "Number of clusters",
     ylab = "SS_within",
     main = "Look for an elbow")
plot(x = ks[-1],
     y = - ssw[-1] + ssw[-length(ssw)], type = "h",
     xlab = "Number of clusters",
     ylab = "Decrement in SS_within",
     main = "Look for a spike")
```

```{r}
km.out.best <- km.out[[3]] 
km.out.best
```

```{r}


```

```{r}

db_cluster_1 <- cbind(db_na, clusterNum = km.out.best$cluster)
db_cluster_1[db_cluster_1$clusterNum == 1,]

hist.data.frame(db_cluster_1[db_cluster_1$clusterNum == 1,])
hist.data.frame(db_cluster_1[db_cluster_1$clusterNum == 2,])
hist.data.frame(db_cluster_1[db_cluster_1$clusterNum == 3,])
hist.data.frame(db_cluster_1[db_cluster_1$clusterNum == 4,])


cluster_1_centr <- t(km.out.best$centers)
colnames(cluster_1_centr) <- make.names(colnames(cluster_1_centr))
cluster_1_centr

pca <- prcomp(var_corr_pos, scale. = T, center = T)
summary(pca)

library(ggfortify)
pca.fortify <- fortify(pca)

pca3.dat <- cbind(pca.fortify, group=km.out.best$cluster)

library(ggplot2)
gg2 <- ggplot(pca3.dat) +
  geom_point(aes(x=PC1, y=PC2, col=factor(group), text=rownames(pca3.dat)), size=2) +
  labs(title = "Visualizing K-Means Clusters Against First Two Principal Components") +
  scale_color_brewer(name="", palette = "Set1")

plot(gg2)
  
theta <- seq(0,2*pi,length.out = 100)
circle <- data.frame(x = cos(theta), y = sin(theta))
p <- ggplot(circle,aes(x,y)) + geom_path()

loadings <- data.frame(pca$rotation, 
                       .names = row.names(pca$rotation))
p + geom_text(data=loadings, 
              mapping=aes(x = PC1, y = PC2, label = .names, colour = .names)) +
  coord_fixed(ratio=1) +
  labs(x = "PC1", y = "PC2", cex=10)


```

#------------------------------------------------------------------------------------------------------

# <span style="color:DODGERBLUE"> Cluster 1- Variabili correlate negativamente</span>

```{r}
var_corr_neg <- data_scale[, c(10,1, 14)]
```


```{r}
var_corr_neg 
```


```{r}
library(cluster)
library(dplyr)

km.out <- list()
sil.out <- list()
x <- vector()
y <- vector()
minClust <- 2     
maxClust <- 10



for (centr in minClust:maxClust) {
  i <- centr-(minClust-1) 
  km.out[i] <- list(kmeans(var_corr_neg, centers = centr, nstart = 50,iter.max = 50))
  sil.out[i] <- list(silhouette(km.out[[i]][[1]], dist(var_corr_neg)))
  x[i] = centr  # value of k
  y[i] = summary(sil.out[[i]])[[4]]  # Silhouette average width
}


library(ggplot2)
ggplot(data = data.frame(x, y), aes(x, y)) + 
  geom_point(size=3) + 
  geom_line() +
  xlab("Number of Cluster Centers") +
  ylab("Silhouette Average Width") +
  ggtitle("Silhouette Average Width as Cluster Center Varies")

ks <- 2:10 # number of clusters we want to try
ssw <- numeric(length(ks)) # vector for the ss_within
for (i in seq_along(ks)) {
  ssw[i] <- kmeans(var_corr_neg, ks[i],iter.max = 50)$tot.withinss
}

plot(x = ks, y = ssw, type = "l",
     xlab = "Number of clusters",
     ylab = "SS_within",
     main = "Look for an elbow")
plot(x = ks[-1],
     y = - ssw[-1] + ssw[-length(ssw)], type = "h",
     xlab = "Number of clusters",
     ylab = "Decrement in SS_within",
     main = "Look for a spike")
```

```{r}
km.out.best <- km.out[[2]] 
km.out.best
```


```{r}

db_cluster_2 <- cbind(db_na, clusterNum = km.out.best$cluster)
db_cluster_2[db_cluster_2$clusterNum == 1,]

hist.data.frame(db_cluster_2[db_cluster_2$clusterNum == 1,])
hist.data.frame(db_cluster_2[db_cluster_2$clusterNum == 2,])
hist.data.frame(db_cluster_2[db_cluster_2$clusterNum == 3,])


cluster_2_centr <- t(km.out.best$centers)
colnames(cluster_2_centr) <- make.names(colnames(cluster_2_centr))
cluster_2_centr

pca <- prcomp(var_corr_neg, scale. = T, center = T)
summary(pca)

library(ggfortify)
pca.fortify <- fortify(pca)

pca3.dat <- cbind(pca.fortify, group=km.out.best$cluster)

library(ggplot2)
gg2 <- ggplot(pca3.dat) +
  geom_point(aes(x=PC1, y=PC2, col=factor(group), text=rownames(pca3.dat)), size=2) +
  labs(title = "Visualizing K-Means Clusters Against First Two Principal Components") +
  scale_color_brewer(name="", palette = "Set1")

plot(gg2)
  
theta <- seq(0,2*pi,length.out = 100)
circle <- data.frame(x = cos(theta), y = sin(theta))
p <- ggplot(circle,aes(x,y)) + geom_path()

loadings <- data.frame(pca$rotation, 
                       .names = row.names(pca$rotation))
p + geom_text(data=loadings, 
              mapping=aes(x = PC1, y = PC2, label = .names, colour = .names)) +
  coord_fixed(ratio=1) +
  labs(x = "PC1", y = "PC2", cex=10)


```
#------------------------------------------------------------------------------------------------------

# <span style="color:DODGERBLUE"> Cluster 2- Variabili bisogni personali</span>

```{r}
var_pers <- data_scale[, c(11,12,13,14)]
```

```{r}
var_pers 
```

```{r}
library(cluster)
library(dplyr)

km.out <- list()
sil.out <- list()
x <- vector()
y <- vector()
minClust <- 2     
maxClust <- 10



for (centr in minClust:maxClust) {
  i <- centr-(minClust-1) 
  km.out[i] <- list(kmeans(var_pers, centers = centr, nstart = 50,iter.max = 50))
  sil.out[i] <- list(silhouette(km.out[[i]][[1]], dist(var_pers)))
  x[i] = centr  # value of k
  y[i] = summary(sil.out[[i]])[[4]]  # Silhouette average width
}


library(ggplot2)
ggplot(data = data.frame(x, y), aes(x, y)) + 
  geom_point(size=3) + 
  geom_line() +
  xlab("Number of Cluster Centers") +
  ylab("Silhouette Average Width") +
  ggtitle("Silhouette Average Width as Cluster Center Varies")

ks <- 2:10 # number of clusters we want to try
ssw <- numeric(length(ks)) # vector for the ss_within
for (i in seq_along(ks)) {
  ssw[i] <- kmeans(var_pers, ks[i],iter.max = 50)$tot.withinss
}

plot(x = ks, y = ssw, type = "l",
     xlab = "Number of clusters",
     ylab = "SS_within",
     main = "Look for an elbow")
plot(x = ks[-1],
     y = - ssw[-1] + ssw[-length(ssw)], type = "h",
     xlab = "Number of clusters",
     ylab = "Decrement in SS_within",
     main = "Look for a spike")
```


```{r}
km.out.best <- km.out[[2]] 
km.out.best
```

```{r}

db_cluster_3 <- cbind(db_na, clusterNum = km.out.best$cluster)
db_cluster_3[db_cluster_3$clusterNum == 1,]

hist.data.frame(db_cluster_3[db_cluster_3$clusterNum == 1,])
hist.data.frame(db_cluster_3[db_cluster_3$clusterNum == 2,])
hist.data.frame(db_cluster_3[db_cluster_3$clusterNum == 3,])
#£hist.data.frame(db_cluster_3[db_cluster_3$clusterNum == 4,])


cluster_3_centr <- t(km.out.best$centers)
colnames(cluster_3_centr) <- make.names(colnames(cluster_3_centr))
cluster_3_centr

pca <- prcomp(var_pers, scale. = T, center = T)
summary(pca)

library(ggfortify)
pca.fortify <- fortify(pca)

pca3.dat <- cbind(pca.fortify, group=km.out.best$cluster)

library(ggplot2)
gg2 <- ggplot(pca3.dat) +
  geom_point(aes(x=PC1, y=PC2, col=factor(group), text=rownames(pca3.dat)), size=2) +
  labs(title = "Visualizing K-Means Clusters Against First Two Principal Components") +
  scale_color_brewer(name="", palette = "Set1")

plot(gg2)
  
theta <- seq(0,2*pi,length.out = 100)
circle <- data.frame(x = cos(theta), y = sin(theta))
p <- ggplot(circle,aes(x,y)) + geom_path()

loadings <- data.frame(pca$rotation, 
                       .names = row.names(pca$rotation))
p + geom_text(data=loadings, 
              mapping=aes(x = PC1, y = PC2, label = .names, colour = .names)) +
  coord_fixed(ratio=1) +
  labs(x = "PC1", y = "PC2", cex=10)


```

#------------------------------------------------------------------------------------------------------

# <span style="color:DODGERBLUE"> Cluster 3 - Variabili caratteristiche cliente</span>

```{r}
var_cliente <- data_scale[,c(1,5,6,9)] #1 5 6 9 10
```

```{r}
var_cliente 
```

```{r}
set.seed(123)
library(cluster)
library(dplyr)

km.out <- list()
sil.out <- list()
x <- vector()
y <- vector()
minClust <- 2     
maxClust <- 10



for (centr in minClust:maxClust) {
  i <- centr-(minClust-1) 
  km.out[i] <- list(kmeans(var_cliente, centers = centr, nstart = 50,iter.max = 50))
  sil.out[i] <- list(silhouette(km.out[[i]][[1]], dist(var_cliente)))
  x[i] = centr  # value of k
  y[i] = summary(sil.out[[i]])[[4]]  # Silhouette average width
}


library(ggplot2)
ggplot(data = data.frame(x, y), aes(x, y)) + 
  geom_point(size=3) + 
  geom_line() +
  xlab("Number of Cluster Centers") +
  ylab("Silhouette Average Width") +
  ggtitle("Silhouette Average Width as Cluster Center Varies")

ks <- 2:10 # number of clusters we want to try
ssw <- numeric(length(ks)) # vector for the ss_within
for (i in seq_along(ks)) {
  ssw[i] <- kmeans(var_cliente, ks[i],iter.max = 50)$tot.withinss
}

plot(x = ks, y = ssw, type = "l",
     xlab = "Number of clusters",
     ylab = "SS_within",
     main = "Look for an elbow")
plot(x = ks[-1],
     y = - ssw[-1] + ssw[-length(ssw)], type = "h",
     xlab = "Number of clusters",
     ylab = "Decrement in SS_within",
     main = "Look for a spike")
```


```{r}
km.out.best <- km.out[[3]] 
km.out.best
```

```{r}

db_cluster_4 <- cbind(db_na, clusterNum = km.out.best$cluster)
db_cluster_4[db_cluster_4$clusterNum == 1,]

hist.data.frame(db_cluster_4[db_cluster_4$clusterNum == 1,])
hist.data.frame(db_cluster_4[db_cluster_4$clusterNum == 2,])
hist.data.frame(db_cluster_4[db_cluster_4$clusterNum == 3,])
hist.data.frame(db_cluster_4[db_cluster_4$clusterNum == 4,])


cluster_4_centr <- t(km.out.best$centers)
colnames(cluster_4_centr) <- make.names(colnames(cluster_4_centr))
cluster_4_centr

pca <- prcomp(var_cliente, scale. = T, center = T)
summary(pca)

library(ggfortify)
pca.fortify <- fortify(pca)

pca3.dat <- cbind(pca.fortify, group=km.out.best$cluster)

library(ggplot2)
gg2 <- ggplot(pca3.dat) +
  geom_point(aes(x=PC1, y=PC2, col=factor(group), text=rownames(pca3.dat)), size=2) +
  labs(title = "Visualizing K-Means Clusters Against First Two Principal Components") +
  scale_color_brewer(name="", palette = "Set1")

plot(gg2)
  
theta <- seq(0,2*pi,length.out = 100)
circle <- data.frame(x = cos(theta), y = sin(theta))
p <- ggplot(circle,aes(x,y)) + geom_path()

loadings <- data.frame(pca$rotation, 
                       .names = row.names(pca$rotation))
p + geom_text(data=loadings, 
              mapping=aes(x = PC1, y = PC2, label = .names, colour = .names)) +
  coord_fixed(ratio=1) +
  labs(x = "PC1", y = "PC2", cex=10)


```

#------------------------------------------------------------------------------------------------------

# <span style="color:DODGERBLUE"> Cluster 5 - Variabili composizione portafogli</span>

```{r}
var_port <- data_scale[,c(20,21,22,23,24)] 
```

```{r}
var_port 
```

```{r}
set.seed(123)
library(cluster)
library(dplyr)

km.out <- list()
sil.out <- list()
x <- vector()
y <- vector()
minClust <- 2     
maxClust <- 10



for (centr in minClust:maxClust) {
  i <- centr-(minClust-1) 
  km.out[i] <- list(kmeans(var_port, centers = centr, nstart = 50,iter.max = 50))
  sil.out[i] <- list(silhouette(km.out[[i]][[1]], dist(var_port)))
  x[i] = centr  # value of k
  y[i] = summary(sil.out[[i]])[[4]]  # Silhouette average width
}


library(ggplot2)
ggplot(data = data.frame(x, y), aes(x, y)) + 
  geom_point(size=3) + 
  geom_line() +
  xlab("Number of Cluster Centers") +
  ylab("Silhouette Average Width") +
  ggtitle("Silhouette Average Width as Cluster Center Varies")

ks <- 2:10 # number of clusters we want to try
ssw <- numeric(length(ks)) # vector for the ss_within
for (i in seq_along(ks)) {
  ssw[i] <- kmeans(var_port, ks[i],iter.max = 50)$tot.withinss
}

plot(x = ks, y = ssw, type = "l",
     xlab = "Number of clusters",
     ylab = "SS_within",
     main = "Look for an elbow")
plot(x = ks[-1],
     y = - ssw[-1] + ssw[-length(ssw)], type = "h",
     xlab = "Number of clusters",
     ylab = "Decrement in SS_within",
     main = "Look for a spike")
```


```{r}
km.out.best <- km.out[[3]] 
km.out.best
```

```{r}

db_cluster_5 <- cbind(db_na, clusterNum = km.out.best$cluster)
db_cluster_5[db_cluster_5$clusterNum == 1,]

hist.data.frame(db_cluster_5[db_cluster_5$clusterNum == 1,])
hist.data.frame(db_cluster_5[db_cluster_5$clusterNum == 2,])
hist.data.frame(db_cluster_5[db_cluster_5$clusterNum == 3,])
hist.data.frame(db_cluster_5[db_cluster_5$clusterNum == 4,])


cluster_5_centr <- t(km.out.best$centers)
colnames(cluster_5_centr) <- make.names(colnames(cluster_5_centr))
cluster_5_centr

pca <- prcomp(var_port, scale. = T, center = T)
summary(pca)

library(ggfortify)
pca.fortify <- fortify(pca)

pca3.dat <- cbind(pca.fortify, group=km.out.best$cluster)

library(ggplot2)
gg2 <- ggplot(pca3.dat) +
  geom_point(aes(x=PC1, y=PC2, col=factor(group), text=rownames(pca3.dat)), size=2) +
  labs(title = "Visualizing K-Means Clusters Against First Two Principal Components") +
  scale_color_brewer(name="", palette = "Set1")

plot(gg2)
  
theta <- seq(0,2*pi,length.out = 100)
circle <- data.frame(x = cos(theta), y = sin(theta))
p <- ggplot(circle,aes(x,y)) + geom_path()

loadings <- data.frame(pca$rotation, 
                       .names = row.names(pca$rotation))
p + geom_text(data=loadings, 
              mapping=aes(x = PC1, y = PC2, label = .names, colour = .names)) +
  coord_fixed(ratio=1) +
  labs(x = "PC1", y = "PC2", cex=10)


```











